---
title: "Exponentials Project"
author: "Phil Allen"
date: "19 December 2015"
output: html_document
---

```{r, cache=TRUE, echo=FALSE}
# Set the pseudo-random seed and the constants given in the question.
set.seed(42)
l=0.2
SS=40
nosim=1000
```

# Overview

We investigate the exponential distribution in R and compare it with the Central Limit Theorem, simulating samples of `r SS` exponentials. 

# Simulations

We take $\lambda$ the rate to be `r l`, so that the population mean and standard deviation are both `r 1/l`. We investigate the distribution of averages of `r SS` exponentials, with `r nosim` trialss.

Sample the exponential distribution into a matrix with `r nosim` rows and `r SS` columns. Apply mean across the columns of each row to obtain a vector of means `r SS` exponentials.

```{r, cache=TRUE}
obs <- matrix(rexp(SS*nosim,l), nosim) # nosim rows of SS observations
sims <- apply(obs,1,mean)
```
# Results

## Sample Mean versus Theoretical Mean

The sample mean across the trials is `r round(mean(sims), 4)`, whereas the theoretical mean is `r 1/l`. It is interesting to see how the mean converges as the number of trials increases from 1 to 1000. Form the means of the subsequences of the observed means.

```{r, cache=TRUE}
means <- cumsum(sims)/(1:nosim)
df <- data.frame(x=1:nosim, y=means)
```

In this case, there is marked divergence from the theortical mean (shown by horiztonal line) during the first 100 trials.

```{r, echo=FALSE}
library(ggplot2)
g <- ggplot(df, aes(x = x, y = y)) + geom_line()
g <- g + geom_hline(yintercept = 5)
g <- g + labs(x = "Number of trials", y = "Mean", title = "Convergence to population mean")
g
```

## Sample Variance versus Theoretical Variance

Highlight the variances you are comparing. Include text that explains your understanding of the differences of the variances.
The population variance of the average of 40 indepedent exp(0.2) variables is $\frac{5^2}{40}$, which is `r 5^2/40`. In this case, the sample variance is `r round(var(sims), 4)`


```{r, cache=TRUE}
var(sims)
sampleMean <- mean(sims)
cumsqerr <- cumsum((sims-sampleMean)^2)
vars <- cumsqerr[2:nosim]/(1:(nosim-1))
dfvars <- data.frame(x=(2:nosim), y=vars)
```

And plot it
```{r, echo=FALSE}
library(ggplot2)
g <- ggplot(dfvars, aes(x = x, y = y)) + geom_line()
g <- g + geom_hline(yintercept = 25/40)
g <- g + labs(x = "Number of trials", y = "Variance", title = "Convergence to population variance")
g
```


## Distribution

We can tell the distribution is approximately normly by plotting a density histogram and superimposing the density function of the corresponding normal distribution $N(5,5 / \sqrt{40})$
```{r}
df3 <- data.frame(x=sims)
dnormSims <- function (x) { dnorm(x, mean=5, sd=5/sqrt(40)) }
```
```{r, echo=FALSE}
g <- ggplot(df3, aes(x = x)) + geom_histogram(binwidth=0.2, alpha=0.2, aes(y = ..density..)) 
g <- g + stat_function(fun = dnormSims, size = 2)
g
```

