---
title: 'Statistical inference: exponentials simulation'
author: "Phil Allen"
date: "25 December 2015"
output:
  pdf_document:
    highlight: zenburn
    keep_tex: yes
  html_document:
    theme: united
---
```{r, echo=FALSE}
library("ggplot2")
library("ggthemes")
```

# Overview

We investigate the exponential distribution in R. By simulation, we compare the mean of independent identically variables with the asymptotic result given by the Central Limit Theorem.

We find that that Central Limit Theorem predicts the variance and distribution of the means fairly well.

# Simulations

```{r, cache=TRUE}
# Set the pseudo-random seed and name the constants given in the question.
set.seed(42)
l=0.2
SS=40
nosim=1000
```

We use a rate $\lambda = 0.2$ so that the population mean and standard deviation are both `r 1/l`. We investigate the distribution of averages of `r SS` exponentials, with `r nosim` trials.

We first sample the exponential distribution into a matrix with `r nosim` rows and `r SS` columns. Then apply mean across the columns of each row to obtain a vector of means of `r SS` exponentials of length `r nosim`.

```{r, cache=TRUE}
obs <- matrix(rexp(SS*nosim,l), nosim) # nosim rows of SS observations
sims <- apply(obs,1,mean)
```

# Results

## Sample Mean versus Theoretical Mean

The sample mean across the simulation trials is `r round(mean(sims), 4)`, whereas the population mean is `r 1/l`. It is interesting to see how the sample mean converges to the population mean as the number of trials increases from 1 to 1000.

To demonstrate this, take cumulative means of the subsequences of the sequence of simulations: first value; mean of first and second value: mean of first, second and third value, and so on. The length of subsequence, which is the number of trials, is plotted on the x-axis in the following plot.

```{r, echo=FALSE}
means <- cumsum(sims)/(1:nosim)
df <- data.frame(x=1:nosim, y=means)
g <- ggplot(df, aes(x = x, y = y)) + theme_economist() + geom_line()
g <- g + geom_hline(yintercept = 5, col="red")
g <- g + xlab("Number of trials") +
  ylab("Mean") +
  ggtitle("Convergence to population mean") +
  theme_economist()
g
```

In this case, there is marked divergence from the theortical mean (shown by red horiztonal line) during the first 100 trials.

## Sample Variance versus Theoretical Variance

Applying the formula given in the lectures, the _population_ variance of the average of 40 independent $exp(0.2)$ variables is $\frac{(1/0.2)^2}{40}$, which is `r 5^2/40`. In this simulation, the sample variance across the trials is `r round(var(sims), 3)`.

## Distribution

We can demonstrate that the distribution of the means is approximately normal by plotting a density histogram of the simulation data, and superimposing the density function of the corresponding normal distribution, $N(5,5 / \sqrt{40})$, as shown by the thick black line.

```{r, echo=FALSE}
df3 <- data.frame(x=sims)
dnormSims <- function (x) { dnorm(x, mean=5, sd=5/sqrt(40)) }
g <- ggplot(df3, aes(x = x))
g <- g + geom_histogram(binwidth=0.2, alpha=0.2, aes(y = ..density..)) 
g <- g + stat_function(fun = dnormSims, size = 2)
g <- g + theme_economist() + 
  xlab("Value of mean") +
  ylab("Density") +
  ggtitle(paste0(nosim, " means of ", SS, " simulated independent variables"))
g
```

However, there is some remaining left skew, reflecting the density function of a single exponential variable, which is ${\lambda}e^{-{\lambda}x}$ for $x\ge{0}$.
