---
title: "Tooth Growth Study"
author: "Phil Allen"
date: "24 December 2015"
output: html_document
---

#Overview

We analysed the ToothGrowth data set, which reports an experiment on the effect of vitamin C dosage and supplement type on tooth growth in guinea pigs.

We found ...

#Data

The data is taken from a study reported in 
[_Crampton, E. W. (1947)_ The growth of the odontoblast of the incisor teeth as a criterion of vitamin C intake of the guinea pig. The Journal of Nutrition 33(5): 491–504](http://jn.nutrition.org/content/33/5/491.full.pdf)

It studies the effect of vitamin C dosage and supplement type using guinea pigs, where tooth growth is taken as a measure of healthly development. This was of particular concern to the Canadian Government during the war years because of the difficulty of providing natural sources of vitamin C to the armed forces for a considerable portion of the year.

The data can be found directly in R:

```{r}
library("datasets")
```

So let's have a look

```{r}
str(ToothGrowth)
summary(ToothGrowth)
```

There are `r dim(ToothGrowth)[1]` observations in a between groups experiment varying both the dose of Vitamin C (ascorbic acid) and the form in which the dose is added to the basal diet: as orange juice or directly as ascorbic acid.

```{r,echo=FALSE}
library("ggplot2")
library("ggthemes")
```

```{r, echo=FALSE}
# Growth (length) against dose, faceted by supplement type.
# This function is for labelling the facets.
flabel <- function(var, value){
    value <- as.character(value)
    if (var=="supp") { 
        value[value=="OJ"] <- "Given Orange Juice"
        value[value=="VC"]   <- "Given Ascorbic Acid"
    }
    return(value)
}
g <- ggplot(data=ToothGrowth, aes(x=dose, y=len)) + theme_economist()
g <- g + geom_point()
g <- g + expand_limits(x=0,y=0) + ylab("Length of ontoblasts (microns)") + xlab("Dose of Vitamin C (mg)")
g <- g + stat_summary(geom = "line", fun.y = mean, size = 1, col = "black")
g <- g + facet_grid(. ~ supp, labeller=flabel)
g
```


#Analysis

The following may be missing the point; should go with dosages first?
```{r,cache=TRUE}
OJgroup <- ToothGrowth[ToothGrowth$supp == "OJ",]
VCgroup <- ToothGrowth[ToothGrowth$supp == "VC",]
t.test(len ~ I(relevel(supp, 2)), paired=TRUE, data=ToothGrowth)
```


#Conclusion

